---
title: "HazelScratchPad"
author: "Hazel John"
date: "April 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
require(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
# Install packages we need just once
library(devtools)
devtools::install_github("juliasilge/tidytext")
install.packages("wordcloud")
```

```{r, echo=FALSE}
# Load required libraries
library(ggplot2)
library(gridExtra)
library(readr)
library(lubridate)
library(tidytext)
library(dplyr)
library(tidyr)
library(stringr)
library(reshape2)
library(wordcloud)
theme_set(theme_bw(base_size = 16))
```

# Data Wrangling

## Step 1: Data Download & Cleanup

Our first task was to download the data, clean it up and extract the data we needed. We decided to do the analysis with just the data from 2014, so the final task was the filter out unneeded data.

We also save the final data sets to submit as part of the project.

This is only run once and not evaluated after that since we can read data from the filtered data files directly.


```{r, eval=FALSE}

# Need to set the path for knitr since the scratchpad is in a 
# subdirectory and we are reading from paths relative to main directory 
opts_knit$set(root.dir = normalizePath('../'))

# Create function to write data frame to zipped rds file
# The dataframe is split into smaller files depending on size
writeToDisk <- function(df, path) {
  # get the size of the data frame
  filesz = object.size(df)
  
  # Figure out if it needs to be split, we try to 
  # split into sizes ~ 250MB (before compression)
  numsplits = filesz %/% (150*1024*1024)
  
  # Split into subsets and write to disk in RDS format
  # so that we can preserve attritubes including type
  if (numsplits > 1) {
    # Split the dataframe into "numsplits" subsets
    df_split <- split(df, ntile(df$`_projectid`, numsplits))
    
    cat("Writing", numsplits, "files with prefix", path, "\n")
    
    # Save data to separate rds files
    # Wrap loop inside invisible() since we are not interested in
    # the return values
    invisible(lapply(names(df_split), function(x) {
      
      write_rds(df_split[[x]], paste0(path, x, "of", numsplits, ".rds.gz"),
                compress = "gz")
    }))
  }
  else {
    
    cat("Writing 1 file with prefix", path, "\n")
    write_rds(df, paste0(path, ".rds.gz"), compress = "gz")
  }
}

# Create function that download file of type "kind", removes special 
# characters and loads the data
retrieveData <- function(kind, needs_cleanup) {
  
  # Create the download link
  url <- paste0("https://s3.amazonaws.com/open_data/csv/opendata_",
                  kind, ".zip")
     
  # Create the path to download the file to           
  zipname <- paste0("data/opendata_", kind, ".zip")
  
  # Create the filename
  filename <- paste0("opendata_", kind, ".csv")
  
  cat("Downloading from", url, "...")
  
  # Download the file
  download.file(url, zipname)
  
  # Donations, resources and essays data files needed cleanup with
  # special characters, escaped characters etc. creating read errors.
  # Data cleanup was done using sed as a system call after
  # realizing that using pipe() to run sed from R was slow.
  # NOTE: The sed script was created on MacOS and might not be portable.
  # Tried to run sed inside pipe - scan(pipe(sed_cmd), sep = ",") 
  # but had too many issues with needing to use multiple escaped characters
  # Also tried readlines() followed by gsub() but the performance was poor.
  if (needs_cleanup) {
    # cleanup is needed so unzip, run sed and then read in data
    
    # unzip the file
    unzip(zipname, filename)
    
    # Create a sed command to clean out special characters
    sed_cmd <- paste0("sed -i '' -f ", kind,
                      "_clnup.sed ", filename)
    
    cat("Running data cleanup for", filename, "...")
    
    # Run the sed command
    system(sed_cmd)
    
    cat("Loading", kind, "...")
    
    # Read in the data
    assign(kind, read_csv(filename), envir=globalenv())
    
    # Remove files
    unlink(zipname)
    unlink(filename)
  }
  else {
    cat("Loading", kind, "...")
    
    # cleanup is not needed, so read in data directly
    assign(kind, read_csv(unz(zipname, filename)), envir=globalenv())

    # Remove zip file
    unlink(zipname)
  }
}

# Create the list the type of data files we want to download
types_list = c("projects", "resources", "donations", "essays")
  
# Note which files need cleanup
needs_cleanup = c(FALSE, TRUE, TRUE, TRUE)

# Download files, remove special characters and load data
for (index in seq(1:4)) {
  retrieveData(types_list[index], needs_cleanup[index])
}

# Convert dates to "Date" format
projects <- projects %>%
  mutate(date_posted = as_date(date_posted),
         date_completed = as_date(date_completed),
         date_thank_you_packet_mailed =
           as_date(date_thank_you_packet_mailed),
         date_expiration = as_date(date_expiration))

donations <- donations %>%
  mutate(donation_timestamp = as_date(donation_timestamp))

# Filter out projects that were posted in 2014
projects <- projects %>% filter(year(date_posted) == 2014)

# Select resources, donations and essays associated with
# 
resources <- resources %>%
  semi_join(projects, by = "_projectid")
donations <- donations %>%
  semi_join(projects, by = "_projectid")
essays <- essays %>%
  semi_join(projects, by = "_projectid")

# Save filtered data to disk
writeToDisk(df=projects, path="data/opendata_2014_projects")
writeToDisk(df=resources, path="data/opendata_2014_resources")
writeToDisk(df=donations, path="data/opendata_2014_donations")
writeToDisk(df=essays, path="data/opendata_2014_essays")

# Let us clean all the variables so as to be able to start 
# with a clean slate
rm(projects, resources, donations, essays, 
   types_list, needs_cleanup, retrieveData, writeToDisk)

# Cleanup memory
gc()
```

## Step 2: Data Upload from disk

This is where we would start after the initial data retrieval. We need to load the data from the RDS files in the data folder

```{r, warning=FALSE}
# Need to set the path for knitr since the scratchpad is in a 
# subdirectory and we are reading from paths relative to main directory 
opts_knit$set(root.dir = normalizePath('..'))

# Create function to load data from the rds files containing the
# name "kind".
uploadData <- function(kind) {
  temp <- list.files(path = "./data", 
                    pattern = paste0(".*", kind, ".*rds.gz"),
                    full.names = TRUE)
  # Read in the data
  tables <- lapply(temp, read_rds)
  
  # Combine multiple (or single) dataframes into one and return
  return (bind_rows(tables))
}

# Read in the different data sets
projects <- uploadData("projects")
resources <- uploadData("resources")
donations <- uploadData("donations")
essays <- uploadData("essays")

# dplyr wants us the put column names in back-quotes when the name
# starts with "_". The quotes however causes errors with knitting 
# So remove all the "_" in the column names 
names(projects) <- gsub("^_", "", names(projects))
names(resources) <- gsub("^_", "", names(resources))
names(donations) <- gsub("^_", "", names(donations))
names(essays) <- gsub("^_", "", names(essays))

```

## Data Visualization & Analysis

```{r, eval=FALSE}

myStackedBarChart <-function(colname) {
  # Create percentage of funding status grouped by colname
  data <- projects %>%
    group_by_(colname) %>%
    summarize(completed = mean(funding_status == "completed", na.rm=TRUE),
      expired = mean(funding_status == "expired", na.rm=TRUE),
      reallocated = mean(funding_status == "reallocated", na.rm=TRUE)) %>%
    melt(variable.name = "funding_type", value.name = "funding_pct")
  
  # Created a stacked bar chart
  data %>%
    arrange_(colname, ~desc(funding_type)) %>%
    ggplot(aes(x=get(colname),
               y=funding_pct, 
               fill = funding_type)) +
    geom_bar(stat="identity") +
    xlab(colname) +
    scale_fill_manual(values=c("green", "red", "blue")) +
    theme(legend.text = element_text(size = 7),
          legend.title = element_text(size = 8),
          legend.key.size = unit(0.25, "cm")) +
    coord_flip()
}

# Identify columns we wanted to create to split into goups.
plot_columns <- c("school_state", "poverty_level", "primary_focus_area",
                  "resource_type", "grade_level")
plot_list <- lapply(plot_columns, FUN = myStackedBarChart)

# Create the layout, give extra space for school_state
lmat <- rbind(c(1,2),
              c(1,3),
              c(4,5))
# Layout the plots in one grid
grid.arrange(grobs = plot_list, layout_matrix = lmat,
             top = "Relation b/w different factors and funding")

```

## Try some text mining

```{r, warning=FALSE}

# Tokenize the essay and remove stop words and include only
# all alphabetic words. All words are lower case, so there is
# no need to transform
essays_tokenized <- essays %>%
  select(projectid, teacherid, essay) %>%
  unnest_tokens(essay_words,essay)  %>%
  filter(!essay_words %in% stop_words$word &
           grepl("^[[:alpha:]]*$", essay_words))

# Get the sentiment lexicon from mrc
nrc <- sentiments %>%
    filter(lexicon == "nrc") %>%
    select(word, sentiment)

# Assign sentiments to words
essays_sentiments <- essays_tokenized %>%
    left_join(nrc, by = c("essay_words" = "word"))

# Include the funding_status of the projects
essays_sentiments <- essays_sentiments %>%
  left_join(projects, by = "projectid") %>%
  select(projectid, funding_status, essay_words, sentiment)

# Count the sentiment frequency
essays_sentiment_freq <-
  essays_sentiments %>%
  group_by(funding_status, sentiment) %>%
  summarise(sentiment_freq = n()) %>%
  group_by(funding_status) %>%
  mutate(occurance_pct = sentiment_freq*100/sum(sentiment_freq)) %>%
  ungroup()
  

# Plot the sentiment frequency and seperate by funding status
essays_sentiment_freq %>%
  filter(funding_status != "reallocated") %>%
  ggplot(aes(x=sentiment, y = occurance_pct, fill = sentiment)) +
  geom_bar(stat="identity") +
  facet_grid(~funding_status) +
  theme(text = element_text(size = 8),
        title = element_text(size = 10),
        legend.key.size = unit(0.5, "cm"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle('Sentiment Occurance% in Funding Essays') +
  coord_flip()

# Compute word frequency in essays regardless of sentiment
# or funding_status
essays_word_freq <-
  essays_sentiments %>%
  group_by(essay_words) %>%
  summarise(completed_freq = sum(funding_status == "completed"),
            expired_freq = sum(funding_status == "expired"))

# Plot the top 10 words for both funding status
p1 <- essays_word_freq %>%
  top_n(n=10, wt=completed_freq) %>%
  ggplot(aes(x=reorder(essay_words, completed_freq),
             y = log10(completed_freq))) +
  geom_bar(stat="identity", fill = "blue") +
  ggtitle('Top 10 Words (completed)') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 10)) +
  xlab("essay_words") +
  coord_flip()

p2 <- essays_word_freq %>%
  top_n(n=10, wt=expired_freq) %>%
  ggplot(aes(x=reorder(essay_words, expired_freq), 
             y = log10(expired_freq))) +
  geom_bar(stat="identity",  fill = "green") +
  ggtitle('Top 10 Words (expired)') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10),
        plot.title = element_text(size = 10)) +
  xlab("essay_words") +
  coord_flip()

grid.arrange(p1, p2, nrow=1)

# Students stands out for both "completed" and "expired" projects,
# So create word clouds without it, to see the rest better
essays_word_freq <- essays_word_freq %>%
  filter(essay_words != "students")

# Create word cloud for funded essays
wordcloud(essays_word_freq$essay_words, essays_word_freq$completed_freq,
          min.freq = 10000, max.words=100, random.order=TRUE,
          rot.per=0.35, colors=brewer.pal(8, "Dark2"))


# Create word cloud for expired essays
wordcloud(essays_word_freq$essay_words, essays_word_freq$expired_freq,
          min.freq = 10000, max.words=100, random.order=TRUE,
          rot.per=0.35, colors=brewer.pal(8, "Dark2"))

```

