---
title: "Basic_exploratory_steve"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(caret)
library(tidyr)
library(broom)

#setwd("~/Dropbox/datasciencefinalproject/DonorsChoose/data")

setwd("C:/Users/Steve/Dropbox/datasciencefinalproject/DonorsChoose/data")

projects <- read_csv("opendata_projects_output.csv")

projects <- projects %>% mutate(percent_funded = (total_donations / total_price_including_optional_support))


#test for normalacy of percent funded
hist(projects$percent_funded)
qqnorm(projects$percent_funded)
print("Data is nowhere near normal, looks like logistic analysis of funded vs non-funded make much more sense.")

#poverty level graph
#amount of goal
projects %>% ggplot(aes(x = poverty_level, y = percent_funded)) + geom_boxplot()
#amount of goal jitter plot
projects %>% ggplot(aes(x = poverty_level, y = funding_status)) + geom_jitter(size = .001)

poverty <- projects %>% group_by(poverty_level, funding_status) %>% summarise(total= n()) %>% 
  group_by(poverty_level) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup 


#percent funded
poverty %>% ggplot(aes(poverty_level, percent)) + geom_point(aes(color = funding_status))


#latitude and longitude level graph by amount of goal
projects %>% ggplot(aes(x = school_latitude, y = percent_funded)) + geom_point()
projects %>% ggplot(aes(x = school_longitude, y = percent_funded)) + geom_point()

#state graph
#percent of goal
projects %>% ggplot(aes(x = school_state, y = percent_funded)) + geom_boxplot()
state <- projects %>% group_by(school_state, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_state) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup

state %>% ggplot(aes(school_state, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#school urban vs rual
projects %>% ggplot(aes(x = school_metro, y = percent_funded)) + geom_boxplot()

urban <- projects %>% group_by(school_metro, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_metro) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup

urban %>% ggplot(aes(school_metro, percent)) + geom_point(aes(color = funding_status))

#teacher prefix
projects %>% ggplot(aes(x = teacher_prefix, y = percent_funded)) + geom_boxplot()

prefix <- projects %>% group_by(teacher_prefix, funding_status) %>% summarise(total= n()) %>% 
  group_by(teacher_prefix) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup

prefix %>% ggplot(aes(teacher_prefix, percent)) + geom_point(aes(color = funding_status))

#charter school

projects %>% group_by(school_charter, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_charter) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_charter, percent)) + geom_point(aes(color = funding_status))

#magnent school

projects %>% group_by(school_magnet, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_magnet) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_magnet, percent)) + geom_point(aes(color = funding_status))

#school year round

projects %>% group_by(school_year_round, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_year_round) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_year_round, percent)) + geom_point(aes(color = funding_status))

#new Leaders New Schools affiliated 

projects %>% group_by(school_nlns, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_nlns) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_nlns, percent)) + geom_point(aes(color = funding_status))

#knowledge is power public charter school (KIPP)

projects %>% group_by(school_kipp, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_kipp) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_kipp, percent)) + geom_point(aes(color = funding_status))

#school charter ready promise

projects %>% group_by(school_charter_ready_promise, funding_status) %>% summarise(total= n()) %>% 
  group_by(school_charter_ready_promise) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(school_charter_ready_promise, percent)) + geom_point(aes(color = funding_status))

#teach for america teacher

projects %>% group_by(teacher_teach_for_america, funding_status) %>% summarise(total= n()) %>% 
  group_by(teacher_teach_for_america) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(teacher_teach_for_america, percent)) + geom_point(aes(color = funding_status))

#ny teaching fellow

projects %>% group_by(teacher_ny_teaching_fellow, funding_status) %>% summarise(total= n()) %>% 
  group_by(teacher_ny_teaching_fellow) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(teacher_ny_teaching_fellow, percent)) + geom_point(aes(color = funding_status))


#primary focus subject

projects %>% group_by(primary_focus_subject, funding_status) %>% summarise(total= n()) %>% 
  group_by(primary_focus_subject) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(primary_focus_subject, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#primary focus area

projects %>% group_by(primary_focus_area, funding_status) %>% summarise(total= n()) %>% 
  group_by(primary_focus_area) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(primary_focus_area, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))


#resource type

projects %>% group_by(resource_type, funding_status) %>% summarise(total= n()) %>% 
  group_by(resource_type) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(resource_type, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#grade level

projects %>% group_by(grade_level, funding_status) %>% summarise(total= n()) %>% 
  group_by(grade_level) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(grade_level, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#total price excluding optional support

projects %>% ggplot(aes(x = total_price_excluding_optional_support, y = percent_funded)) + geom_smooth()

#students reached

projects %>% ggplot(aes(x = students_reached, y = percent_funded)) + geom_smooth()

#eligible for double your impact match

projects %>% group_by(eligible_double_your_impact_match, funding_status) %>% summarise(total= n()) %>% 
  group_by(eligible_double_your_impact_match) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(eligible_double_your_impact_match, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#eligible for almost home match

projects %>% group_by(eligible_almost_home_match, funding_status) %>% summarise(total= n()) %>% 
  group_by(eligible_almost_home_match) %>% mutate(totals = sum(total), percent = total/totals) %>% ungroup %>% 
  ggplot(aes(eligible_almost_home_match, percent)) + geom_point(aes(color = funding_status)) + theme(
    axis.text.x=element_text(angle=90, size=8))

#date posted

projects %>% ggplot(aes(x = date_posted, y = percent_funded)) + geom_smooth()

projects %>% mutate(sintry = asin(percent_funded)) %>% ggplot(aes(x = date_posted, y = sintry)) + geom_smooth()


```


```{r eval=FALSE}

##MODELING

#This chunk is just trying a numer of things, see next chunk for best model so far

#create column for binary 1=funded, 0 = not funded regression
projects <- projects %>% mutate(funded = ifelse(funding_status == "completed", 1 , 0))

#split into test and train data sets (90% in train)
n_test <- round(nrow(projects) / 10)
test_indices <- sample(1:nrow(projects), n_test, replace=FALSE)
test <- projects[test_indices,]
train <- projects[-test_indices,]


#root mean square error defined in class to use test models, think this will work with logistic output, not 100% sure
RMSE <- function(true_ratings, predicted_ratings){
    sqrt(mean((true_ratings - predicted_ratings)^2))
}

#using confusion matrix to test accurary

tab <- table(  dat$glm_pred , dat$truth )
conf_matrix <- confusionMatrix(tab)
conf_matrix$table
conf_matrix$overall["Accuracy"]

#model based off of date alone

fit <- glm(funded ~ date_posted, data=train, family = "binomial")


#see train documentation for other logistic methods http://topepo.github.io/caret/Logistic_Regression.html
control <- trainControl(method='cv', number=10, p=.8)

logreg <- train(funded ~ total_price_excluding_optional_support,
             data = train,
             method = "logicBag")

mod_fit <- train(funded ~ total_price_excluding_optional_support + primary_focus_area + school_charter + primary_focus_subject + students_reached,  data=train, method="glm", family="binomial")


#fliter out less used variables
projects2 <- projects %>% filter(school_year_round == 'f'& school_magnet == 'f'& school_charter == 'f'& school_nlns == 'f'& school_kipp == 'f'& school_charter_ready_promise == 'f'& teacher_teach_for_america == 'f'& teacher_ny_teaching_fellow == 'f')


#trying most fitting variables
mod_fit2 <- train(funded ~ total_price_excluding_optional_support + primary_focus_area + primary_focus_subject + students_reached + date_posted + teacher_prefix + school_state + school_metro + resource_type + grade_level + eligible_double_your_impact_match + date_posted,  data=projects2, method="glm", family="binomial")

#picking a few mor likely important variables
mod_fit3 <- train(funded ~ total_price_excluding_optional_support + primary_focus_area + primary_focus_subject + date_posted + teacher_prefix + school_state + eligible_double_your_impact_match + date_posted,  data=projects2, method="glm", family="binomial")

#even fewer variables
mod_fit4 <- train(funded ~ total_price_excluding_optional_support + primary_focus_area + primary_focus_subject + students_reached + date_posted,  data=projects2, method="glm", family="binomial")

mod_fit5 <- train(funded ~ total_price_excluding_optional_support + primary_focus_area + date_posted,  data=projects2, method="glm", family="binomial")

mod_fit6 <- train(funded ~ total_price_excluding_optional_support,  data=projects2, method="glm", family="binomial")
```

```{r}
#modeling start

#split into test and train data sets (90% in train), did not use since train() uses cross correlation itself to calculate RMSE, but could be used in future to double check
n_test <- round(nrow(projects) / 10)
test_indices <- sample(1:nrow(projects), n_test, replace=FALSE)
test <- projects[test_indices,]
train <- projects[-test_indices,]

#fliter out projects that had lesser used variables that seemed to have an effect
projects2 <- projects %>% filter(school_year_round == 'f'& school_magnet == 'f'& school_charter == 'f'& school_nlns == 'f'& school_kipp == 'f'& school_charter_ready_promise == 'f'& teacher_teach_for_america == 'f'& teacher_ny_teaching_fellow == 'f')


#simple model to start with, but best predictor so far, can just add variables total price and retrain. view model to see RMSE determined by cross correlation.
model <- train(funded ~ total_price_excluding_optional_support,  data=projects2, method="glm", family="binomial")

